---
title: "Data Basics"
author: Zach del Rosario
date: 2020-05-03
output: github_document
time: 10
reading: 0
---

# Data: Basics

*Purpose*: When first studying a new dataset, there are very simple checks we
should perform first. These are those checks.

Additionally, we'll have our first look at the *pipe operator*, which will be
super useful for writing code that's readable.

*Reading*: (None)

```{r setup}
library(tidyverse)
```

```{r options, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

## First Checks
<!-- -------------------------------------------------- -->

### __q0__ Run the following chunk:

*Hint*: You can do this either by clicking the green arrow at the top-right of
the chunk, or by using the keybaord shortcut `Shift` + `Cmd/Ctrl` + `Enter`.

```{r }
head(iris)
```

This is a *dataset*; the fundamental object we'll study throughout this course.
Some nomenclature:

- The `1, 2, 3, ...` on the left enumerate the **rows** of the dataset
- The names `Sepal.Length`, `Sepal.Width`, `...` name the **columns** of the dataset
- The column `Sepal.Length` takes **numeric** values
- The column `Species` takes **string** values

### __q1__ Load the `tidyverse` and inspect the `diamonds` dataset. What do the
`cut`, `color`, and `clarity` variables mean?

*Hint*: You can run `?diamonds` to get information on a built-in dataset.

```{r load-packages, eval=FALSE}
?diamonds
## cut measures the quality of the cut (str), color measures the diamond color from best to worst (str), clarity measures how clear diamond is (str)

### __q2__ Run `glimpse(diamonds)`; what variables does `diamonds` have?
glimpse(diamonds)
## variables: carat, cut, color, clarity, depth, table, price, x (length), y (width), z (depth)

### __q3__ Run `summary(diamonds)`; what are the common values for each of the variables? How widely do each of the variables vary?
summary(diamonds)
## carat: 4.81, cut: ideal to fair, color: D to J (best to worst), clarity: SI1 to VVS1 (third worst to third best), depth: 36mm, table: 52, price: 18497, x: 10.74, y: 58.9, z: 31.8 

*Hint*: The `Median` and `Mean` are common values, while `Min` and `Max` give us
a sense of variation.

```{r q3-task}
summary(diamonds)
```

**Observations**:

- Carat: While the max carat value is high at 5.01, most of the diamonds are around .7-.8 carats
- Cut: Most diamonds have a very good or better quality cut, I would assume because they are very careful with cuts of diamonds because they have so much value
- Color: There are relatively few diamonds with a bad color, but most do not have the best color but still good color
- Clarity: As the clarity of the diamond increases, the quantity of the diamonds decrease
- Depth: Doesn't vary much, sits typically in the 60s
- Table: Doesn't vary much, sits typically in the high 50s
- Price: On average, a diamond from this sample costs $3933, and the max is an outlier at $18.8k
- Size (x,y,z): The diamonds are typically around 5mm long and wide and around 3mm deep

You should always analyze your dataset in the simplest way possible, build
hypotheses, and devise more specific analyses to probe those hypotheses. The
`glimpse()` and `summary()` functions are two of the simplest tools we have.

## The Pipe Operator
<!-- -------------------------------------------------- -->

Throughout this class we're going to make heavy use of the *pipe operator*
`%>%`. This handy little function will help us make our code more readable.
Whenever you see `%>%`, you can translate that into the word "then". For
instance

```{r pipe-example}
diamonds %>%
  group_by(cut) %>%
  summarize(carat_mean = mean(carat))
```

Would translate into the tiny "story"

- Take the `diamonds` dataset, *then*
- Group it by the variable `cut`, *then*
- summarize it by computing the `mean` of `carat`

*What the pipe actually does*. The pipe operator `LHS %>% RHS` takes its
left-hand side (LHS) and inserts it as an the first argument to the function on
its right-hand side (RHS). So the pipe will let us take `glimpse(diamonds)` and
turn it into `diamonds %>% glimpse()`.

### __q4__ Use the pipe operator to re-write `summary(diamonds)`.

```{r q4-task}
diamonds %>% summary()

## Reading Data
<!-- -------------------------------------------------- -->

So far we've only been looking at built-in datasets. Ultimately, we'll want to read in our own data. We'll get to the art of loading and *wrangling* data later, but for now, know that the `readr` package provides us tools to read data. Let's quickly practice loading data below.

### __q5__ Use the function `read_csv()` to load the file `"./data/tiny.csv"`.

```{r q5-task}
read_csv("./data/tiny.csv")
df_q5
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-data00-basics-assignment.Rmd).
