---
title: "Vis: Bar Charts"
author: Zach del Rosario
date: 2020-05-13
output: github_document
time: 30
reading: 30
---

# Vis: Bar Charts

*Purpose*: *Bar charts* are a key tool for EDA. In this exercise, we'll learn
how to construct a variety of different bar charts, as well as when---and when
*not*---to use various charts.

*Reading*: [Bar Charts](https://rstudio.cloud/learn/primers/3.2)
*Topics*: (All topics)
*Reading Time*: ~30 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

### __q1__ In the reading, you learned the relation between `geom_bar()` and
`geom_col()`. Use that knowledge to convert the following `geom_bar()` plot into
the same visual using `geom_col()`.

```{r q1-task}
mpg %>%
  ggplot(aes(trans)) +
  geom_bar()

mpg %>% 
  count(trans) %>% 
  ggplot(mapping = aes(x = trans, y = n)) +
  geom_col()
```

The reading mentioned that when using `geom_col()` our x-y data should be
`1-to-1`. This next exercise will probe what happens when our data are not
`1-to-1`, and yet we use a `geom_col()`. Note that a
[one-to-one](https://en.wikipedia.org/wiki/Injective_function) function is one
where each input leads to a single output. For the `mpg` dataset, we can see
that the pairs `cty, hwy` clearly don't have this one-to-one property:

```{r mpg-not-1-to-1}
## NOTE: Run this chunk for an illustration
mpg %>% filter(cty == 20)
```

### __q2__ The following code attempts to visualize `cty, hwy` from `mpg` using
`geom_col()`. There's something fishy about the `hwy` values; what's wrong here?

*Hint*: Try changing the `position` parameter for `geom_col()`.

```{r q2-task}
mpg %>%
  ggplot(aes(x = cty, y = hwy), position = "stacked") +
  geom_col(color="white")
```

**Observations**:
- This graph does a poor job of displaying the data effectively. We are using a column graph for data that isn't one-to-one. This raises issues because one value in cty can have many values for hwy. This has made it so the data is confusing to read on this graph and derive any sort of meaning from because we can't really see the hwy values or the amount that there are. 

A more standard way to visualize this kind of data is a *scatterplot*, which
we'll study later. For now, here's an example of a more effective way to
visualize `cty` vs `hwy`:

```{r q2-compare}
## NOTE: Run this chunk for an illustration
mpg %>%
  ggplot(aes(cty, hwy)) +
  geom_point()
```

### __q3__ The following are two different visualizations of the `mpg` dataset.
Document your observations between the `v1` and `v2` visuals. Then, determine
which---`v1` or `v2`---enabled you to make more observations. What was the
difference between the two visuals?

```{r q4-v1}
## TODO: Run this code without changing, describe your observations on the data
mpg %>%
  ggplot(aes(class, fill = class)) +
  geom_bar()
```

**Observations**:
- This graph clearly displays the amount of cars in each class of car. From this graph, we can see that SUVs are the most common class of car, followed by compact cars. 2seater cars are by far the least common, followed by minivans. This graph is easily readable and gives us straight-forward information: class of car and count in dataset. 

```{r q4-v2}
## TODO: Run this code without changing, describe your observations on the data
mpg %>%
  ggplot(aes(class, fill = drv)) +
  geom_bar()
```

**Observations**:
- This graph is fairly easy to read, showing the class of car, the count in the data set, and the type of drive that the car has. From the height of the bars, we see the count of each class, and from the color blocks of the bars, we see the distribution of type of drive for the car. For example, all 2 seaters are rear wheel drive while subcompact cars are mostly front wheel drive, followed by rear wheel drive, then four wheel drive.

**Compare `v1` and `v2`**:

- Which visualization---`v1` or `v2`---enabled you to make more observations?
  The v2 visualization enabled me to make more observations because there was an additional variable that was displayed. With the v1 visualization, I could only see the count per class of car, while with the v2 visualization, I can see the count per class of car and the distribution of type of drive per class. 
  
- What was the difference between `v1` and `v2`?
  The v1 visualization displays the relationship between two variables while the v2 visualizations displays the relationship between three variables. This made it so information could be derived from v2 while the v1 visualization was more visually simple and easy to comprehend at a glance. 

### __q4__ The following code has a bug; it does not do what its author intended.
Identify and fix the bug. What does the resulting graph tell you about the
relation between `manufacturer` and `class` of cars in this dataset?

*Note*: I use a `theme()` call to rotate the x-axis labels. We'll learn how to
do this in a future exercise.

```{r q4-task}
mpg %>%
  ggplot(aes(x = manufacturer, fill = class)) +
  geom_bar(position = "dodge") +
  theme(axis.text.x = element_text(angle = 270, vjust = 0.5, hjust = 0))
```

**Observations**
- The issues with the initial graph was that is was not doing position = "dodge" as the author had intended. To get the graph to display as intended, the statement needed to be nested in geom_bar and not an argument in aes(). This graph allows us to see the distribution of amount of specific class of car by manufacturer. Toyota produces the widest distribution of classes of cars, while some manufacturers only produce one type of car. This graph can be hard to read when not looking at specific subsections but does provide interesting information when examined. 

### __q5__ The following graph is hard to read. What other form of faceting would
make the visual more convenient to read? Modify the code below with your
suggested improvement.

```{r q5-task}
mpg %>%
  ggplot(aes(x = cyl)) +
  geom_bar() +
  facet_wrap(. ~ manufacturer)
```

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis01-bar-charts-assignment.Rmd).
