---
title: "Vis: Boxplots and Counts"
author: Zach del Rosario
date: 2020-06-04
output: github_document
dep: e-data02-derive
time: 20
reading: 20
---

# Vis: Boxplots and Counts

*Purpose*: *Boxplots* are a key tool for EDA. Like histograms, boxplots give us a sense of "shape" for a distribution. However, a boxplot is a *careful summary* of shape. This helps us pick out key features of a distribution, and enables easier comparison of different distributions.

*Reading*: [Boxplots and Counts](https://rstudio.cloud/learn/primers/3.4)
*Topics*: (All topics)
*Reading Time*: ~20 minutes

```{r setup, include=FALSE}
# knitr options
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
library(tidyverse)

```

A *subtle point* from the primer is that we can use `dplyr` to generate new
facts about our data, then use `ggplot2` to visualize those facts. We'll
practice this idea in a number of ways.

### __q1__ Use a `cut_*` verb to create a categorical variable out of `carat`. Tweak
the settings in your cut and document your observations.

*Hint*: Recall that we learned about `cut_interval, cut_number, cut_width`. Take
your pick!

```{r q1-task}
diamonds %>%
  ## TODO: Use mutate to add the column `carat_cut`
  mutate(carat_cut = cut_width(carat, width = 1)) %>%
  ggplot(aes(x = carat_cut, y = price)) +
  geom_boxplot()
```

**Observations**
- In general, the median price increases with the carat value. With a carat value 0.5 or lower, the median price is around 1000 dollars and the most expensive diamond of that group reaches around 350 dollars. On the other hand, the median price of a diamond between 3.5 and 4.5 carats is over 15000 dollars. For the most part, the data is centered, at least in the middle 50% of the data. For the low and high end of the carat values, the IQR is tiny, whereas with more common carat values in the middle, the IQR is relatively large. As carat values increase, the top whisker gets shorter because it reaches the limit of the most expensive diamond. Diamonds with carat values between 0.5 and 1.5 have a wide range with many outliers on the upper end.

### __q2__ The following code visualizes the count of diamonds of *all* carats according to their cut and color. Modify the code to consider *only* diamonds with `carat >= 2`. Does the most common group of cut and color change?

```{r q2-ref}
## NOTE: No need to modify; run and inspect
diamonds %>%
  count(cut, color) %>%
  ggplot(aes(cut, color, fill = n)) +
  geom_tile()
```

Modify the following code:

```{r q2-task}
## TODO: Modify the code below to consider only diamonds with carat >= 2
diamonds %>%
  filter(carat >= 2) %>%
  count(cut, color) %>%
  ggplot(aes(cut, color, fill = n)) +
  geom_tile()
```

**Observations**:
- Did the most common group of cut and color change?
  Yes, after filtering for carat values greater than or equal to 2, the most common group switched from ideal cut with G color to premium cut with I color.

### __q3__ The following plot has overlapping x-axis labels. Use a verb from the reading to `flip` the coordinates and improve readability.

```{r q3-task}
mpg %>%
  ggplot(aes(manufacturer, hwy)) +
  geom_boxplot() +
  coord_flip()
```

This is a simple---but important---trick to remember when visualizing data with many categories.

<!-- include-exit-ticket -->
# Exit Ticket
<!-- -------------------------------------------------- -->

Once you have completed this exercise, make sure to fill out the **exit ticket survey**, [linked here](https://docs.google.com/forms/d/e/1FAIpQLSeuq2LFIwWcm05e8-JU84A3irdEL7JkXhMq5Xtoalib36LFHw/viewform?usp=pp_url&entry.693978880=e-vis03-boxplots-assignment.Rmd).
