---
title: "RMS Titanic"
author: "Lily Dao"
date: 2023-01-28
output:
  github_document:
    toc: true
prerequisites:
  - e-vis01-bar-charts
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Needs Improvement | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations, or observations incorrect | Documented correct observations based on analysis |
| Supported | Some observations not clearly supported by analysis | All observations clearly supported by analysis (table, graph, etc.) |
| Assessed | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support |
| Specified | Uses the phrase "more data are necessary" without clarification | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

### __q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

- Variables for this dataset include Class, Sex, Age, Survived, and n for this dataset.

### __q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of souls aboard. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- Are there any differences?
  - The Wikipedia article estimates that there were 2,224 passengers and crew aboard whereas this code calculates 2,201 people on the Titanic at the time of the sinking. This is a 23 passenger difference. 
- If yes, what might account for those differences?
  - This difference may be due to the way that data was transferred or collected between the two sources. For this dataset in R, the children are divided by sex whereas on Wikipedia they are not. This could lead to miscounting, double counting, or leaving someone out. I assume the Titanic have some way of tracking the passengers on the ship due to the extensive data of what class each of the passengers was. This would leave me to believe that the crew was the least documented of the people on the boat, which could lead to some miscounting on either end.

### __q3__ Create a plot showing the count of persons who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>% 
  group_by(Class, Sex) %>% 
  filter(Survived == "Yes") %>% 
  summarise(total = sum(n)) %>% 
  ggplot() +
  geom_col(mapping = aes(x = Class, y = total, fill = Sex), position = "dodge")
```

**Observations**:

- From this graph, I notice that the male crew members were the largest group of people to survive, with female crew members being the smallest group of people to survive. This could be skewed if there were significantly less female crew members on board to begin with. Another thing worth noting is that 2nd and 3rd class female passengers had about the same number who survived, while 1st class had significantly more. For male passengers, 2nd class had the least survivors and 3rd class had the most. It is difficult to make sense of any correlation between class, sex, and survival without being able to see how many people survived relative to how many people were on board.

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop
```

### __q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3. Is there anything *fishy* in your plot?

```{r q4-task}
df_titanic %>% 
  group_by(Class, Sex) %>% 
  filter(Survived == "Yes") %>% 
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge")
```

**Observations**:

- This graph explains a lot of the strange things we saw in the previous graph of some groups having a large group of survivors with other groups having a small group of survivors. An important thing to consider is how many total people in each group were there to compare with how many survived. This graph shows that males and females from the crew had similar survival rates and nearly 100% survival. Similarly, the males and females of the 3rd class passengers had similar survival rates, but their survival rate was just under 80%. With 2nd class passengers, we see the greatest difference in survival rates, with male survival rate at around 55% and female survival rate just over 85%. Generally, if the survival rates between sexes differs significantly, it is the females that have the greater survival rate. 

- Is there anything *fishy* going on in your plot?
  - Something strange with this plot is that the survival rates are all relatively high. Knowing that around 2/3 of those on the ship died, these statistics seem off, especially considering 2nd class male passengers is the only group with a lower survival rate than 67%.

### __q5__ Create a plot showing the group-proportion of occupants who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_titanic %>% 
  group_by(Sex, Age, Class) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge") +
  facet_grid(. ~ Age)
```

**Observations**:

- This graph looks significantly different than the graph I made earlier in this challenge. A common trend in this data is that men have much lower survival rates than women. 3rd class passengers, in both age and sex groups, have less than a 50 percent survival rate. Overall, males make up for more of the fatality rate than females do as there were more males on board and a greater percentage that did not survive the sinking. Most children survived, but there were relatively few children on the ship to begin with. This higher survival rate of women and children could be associated with the expectation of men to protect their wives and families in the event of danger.
- If you saw something *fishy* in q4 above, use your new plot to explain the fishy-ness.
  - This new plot is more accurate than the last plot and that is due to how the code was written. In the former graph, the code filtered those who had survived, then calculated the proportion. This led to a graph showing the proportion of adults who had survived out of the total people in that group (class, sex). For example, the previous graph said that 57/62 1st class men survived, but really this meant 57/62 first class adult men survived out of all first class men on the ship. This led to a wildly inaccurate graph. On the other hand, the latter graph filtered after calculating the proportion of survivors. This meant that the proportion of survivors was out of all people from that respective group on the ship, rather than all who survived. This gives a proper representation of the data and how many people survived the titanic sinking. 

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
