---
title: "Michelson Speed-of-light Measurements"
author: "Lily Dao"
date: 2022-02-06
output:
  github_document:
    toc: true
prerequisites:
  - e-data02-derive
---

*Purpose*: When studying physical problems, there is an important distinction between *error* and *uncertainty*. The primary purpose of this challenge is to dip our toes into these factors by analyzing a real dataset.

*Reading*: [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (Optional)

<!-- include-rubric -->

# Grading Rubric

<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual

<!-- ------------------------- -->

| Category    | Needs Improvement                                                                                                | Satisfactory                                                                                                               |
|------------------|-----------------------------|-------------------------|
| Effort      | Some task **q**'s left unattempted                                                                               | All task **q**'s attempted                                                                                                 |
| Observed    | Did not document observations, or observations incorrect                                                         | Documented correct observations based on analysis                                                                          |
| Supported   | Some observations not clearly supported by analysis                                                              | All observations clearly supported by analysis (table, graph, etc.)                                                        |
| Assessed    | Observations include claims not supported by the data, or reflect a level of certainty not warranted by the data | Observations are appropriately qualified by the quality & relevance of the data and (in)conclusiveness of the support      |
| Specified   | Uses the phrase "more data are necessary" without clarification                                                  | Any statement that "more data are necessary" specifies which *specific* data are needed to answer what *specific* question |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability                                 | Code sufficiently close to the [style guide](https://style.tidyverse.org/)                                                 |

## Due Date

<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due **at midnight** before the day of the class discussion of the challenge. See the [Syllabus](https://docs.google.com/document/d/1qeP6DUS8Djq_A0HMllMqsSqX3a9dbcx1/edit?usp=sharing&ouid=110386251748498665069&rtpof=true&sd=true) for more information.

```{r message=FALSE, warning=FALSE}
# Libraries
library(tidyverse)
library(googlesheets4)

url <- "https://docs.google.com/spreadsheets/d/1av_SXn4j0-4Rk0mQFik3LLr-uf0YdA06i3ugE6n-Zdo/edit?usp=sharing"

# Parameters
LIGHTSPEED_VACUUM    <- 299792.458 # Exact speed of light in a vacuum (km / s)
LIGHTSPEED_MICHELSON <- 299944.00  # Michelson's speed estimate (km / s)
LIGHTSPEED_PM        <- 51         # Michelson error estimate (km / s)
```

*Background*: In 1879 Albert Michelson led an experimental campaign to measure the speed of light. His approach was a development upon the method of Foucault[3], and resulted in a new estimate of $v_0 = 299944 \pm 51$ kilometers per second (in a vacuum). This is very close to the modern *exact* value of `r LIGHTSPEED_VACUUM`. In this challenge, you will analyze Michelson's original data, and explore some of the factors associated with his experiment.

I've already copied Michelson's data from his 1880 publication; the code chunk below will load these data from a public googlesheet.

*Aside*: The speed of light is *exact* (there is **zero error** in the value `LIGHTSPEED_VACUUM`) because the meter is actually [*defined*](https://en.wikipedia.org/wiki/Metre#Speed_of_light_definition) in terms of the speed of light!

```{r read-sheet}
## Note: No need to edit this chunk!
gs4_deauth()
ss <- gs4_get(url)
df_michelson <-
  read_sheet(ss) %>%
  select(Date, Distinctness, Temp, Velocity) %>%
  mutate(Distinctness = as_factor(Distinctness))

df_michelson %>% glimpse
```

*Data dictionary*:

-   `Date`: Date of measurement
-   `Distinctness`: Distinctness of measured images: 3 = good, 2 = fair, 1 = poor
-   `Temp`: Ambient temperature (Fahrenheit)
-   `Velocity`: Measured speed of light (km / s)

### **q1** Re-create the following table (from Michelson (1880), pg. 139) using `df_michelson` and `dplyr`. Note that your values *will not* match those of Michelson *exactly*; why might this be?

| Distinctness | n   | MeanVelocity |
|--------------|-----|--------------|
| 3            | 46  | 299860       |
| 2            | 39  | 299860       |
| 1            | 15  | 299810       |

```{r q1-task}
## TODO: Compute summaries
df_q1 <- 
df_michelson %>%
  group_by(Distinctness) %>%
  summarise(MeanVelocity = mean(Velocity), n = n())
df_q1 %>%
  arrange(desc(Distinctness)) %>%
  knitr::kable()
```

**Observations**: - Write your observations here! - Most of the measurements were rated good distinctness, but this was followed closely by fair distinctness. 15% of the measurements were rated to have poor distinctness and this group of measurements had a mean velocity of 299808.0 km/s, over 50 km/s below the other two groups. The groups of good and fair distinctness have relatively close mean velocities that sit around 299860 km/s. - Why might your table differ from Michelson's? - My table is different from Michelson's because his is rounded to the nearest tens place, whereas mine is rounded to the nearest tenths place. That is why we have similar values in the same range but not the same values in our tables for MeanVelocity.

The `Velocity` values in the dataset are the speed of light *in air*; Michelson introduced a couple of adjustments to estimate the speed of light in a vacuum. In total, he added $+92$ km/s to his mean estimate for `VelocityVacuum` (from Michelson (1880), pg. 141). While the following isn't fully rigorous ($+92$ km/s is based on the mean temperature), we'll simply apply this correction to all the observations in the dataset.

### **q2** Create a new variable `VelocityVacuum` with the $+92$ km/s adjustment to `Velocity`. Assign this new dataframe to `df_q2`.

```{r q2-task}
## TODO: Adjust the data, assign to df_q2
df_q2 <-
df_michelson %>%
  mutate(VelocityVacuum = Velocity + 92)
df_q2
```

As part of his study, Michelson assessed the various potential sources of error, and provided his best-guess for the error in his speed-of-light estimate. These values are provided in `LIGHTSPEED_MICHELSON`---his nominal estimate---and `LIGHTSPEED_PM`---plus/minus bounds on his estimate. Put differently, Michelson believed the true value of the speed-of-light probably lay between `LIGHTSPEED_MICHELSON - LIGHTSPEED_PM` and `LIGHTSPEED_MICHELSON + LIGHTSPEED_PM`.

Let's introduce some terminology:[2]

-   **Error** is the difference between a true value and an estimate of that value; for instance `LIGHTSPEED_VACUUM - LIGHTSPEED_MICHELSON`.
-   **Uncertainty** is an analyst's *assessment* of the error.

Since a "true" value is often not known in practice, one generally does not know the error. The best they can do is quantify their degree of uncertainty. We will learn some means of quantifying uncertainty in this class, but for many real problems uncertainty includes some amount of human judgment.[2]

### **q3** Compare Michelson's speed of light estimate against the modern speed of light value. Is Michelson's estimate of the error (his uncertainty) greater or less than the true error?

```{r q3-task}
## TODO: Compare Michelson's estimate and error against the true value
print(paste(RealError = LIGHTSPEED_MICHELSON - LIGHTSPEED_VACUUM, EstimateError = LIGHTSPEED_PM))

```

**Observations**: - Is Michelson's estimate of the error (his uncertainty) greater or less than the true error? - Michelson's estimate of the error is less than the true error. The estimate of his error was plus or minus 51 km/s, but the actual error was 151.542 km/s which is relatively far off. - Make a quantitative comparison between Michelson's uncertainty and his error. - Michelson's uncertainty was about a third of this actual error, with his uncertainty being 51 km/s and his error being 151.542 km/s. This means that his error was about three times greater than he anticipated through his measurements.

The following plot shows all of Michelson's data as a [control chart](https://en.wikipedia.org/wiki/Control_chart); this sort of plot is common in manufacturing, where it is used to help determine if a manufacturing process is under [statistical control](https://en.wikipedia.org/wiki/Statistical_process_control). Each dot is one of Michelson's measurements, and the grey line connects the mean taken for each day. The same plot also shows simulated data using a probability model. We'll get into statistics later in the course; for now, let's focus on understanding what real and simulated data tend to look like.

### **q4** Inspect the following plot with the `Real` Michelson data and `Simulated` data from a probability model. Document the similarities and differences between the data under *observe* below.

```{r q4-cf-real-simulated}
## Note: No need to edit this chunk!
## Calibrate simulated data
v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Date, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_line(
    data = . %>%
      group_by(Date, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Date of Measurement (1879)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**: Similarities

\- Generally, both graphs follow a rising and falling pattern over time so the simulation and reality lined up when looking at velocity in vacuum increases and decreases.

\- Both graphs have about the same amount of the graph going out of the uncertainty range (1/3).

\- The mean velocity in vacuum tends to stay generally within 100 km/s of the actual speed of light, with about two exceptions in simulated data and six exceptions in the real data.

Differences

\- The simulated results tend to stay within the estimated error a lot more than the actual results do. The simulated results have two points outside of the estimated error while the real results have six points outside of the estimated error. This would makes sense because if Michelson simulated this and found that his data sat way out of his error range, he would have probably adjusted his uncertainty.

\- The real results vary more than the simulated results, with the simulated results generally staying relatively flat and experiencing less rises and falls.

\- The real results hopped out of the uncertainty range early June whereas the simulated results hopped out of the uncertainty range in late June.

### **q5** You have access to a few other variables. Construct a few visualizations of `VelocityVacuum` against these other factors. Are there other patterns in the data that might help explain the difference between Michelson's estimate and `LIGHTSPEED_VACUUM`?

```{r}
#VacuumVelocity vs. Distinctness

v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Distinctness, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_boxplot(
    data = . %>%
      group_by(Distinctness, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity),
    size = 0.8
  ) +

  facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Distinctness",
    y = "Velocity (in Vacuum)"
  )
##VacuumVelocity vs. Temp

v_mean <-
  df_q2 %>%
  summarize(m = mean(VelocityVacuum)) %>%
  pull(m)
v_sd <-
  df_q2 %>%
  summarize(s = sd(VelocityVacuum)) %>%
  pull(s)

## Visualize
set.seed(101)
df_q2 %>%
  mutate(Simulated = rnorm(n(), mean = v_mean, sd = v_sd)) %>%
  rename(Real = VelocityVacuum) %>%
  pivot_longer(
    cols = c(Simulated, Real),
    names_to = "source",
    values_to = "velocity"
  ) %>%

  ggplot(aes(Temp, velocity)) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON,
    linetype = "dotted"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON - LIGHTSPEED_PM,
    linetype = "dashed"
  ) +
  geom_hline(
    yintercept = LIGHTSPEED_MICHELSON + LIGHTSPEED_PM,
    linetype = "dashed"
  ) +

  geom_line(
    data = . %>%
      group_by(Temp, source) %>%
      summarize(velocity_mean = mean(velocity)),
    mapping = aes(y = velocity_mean),
    color = "grey50"
  ) +
  geom_point(
    mapping = aes(y = velocity, color = Distinctness),
    size = 1.0
  ) +

  facet_grid(source~.) +
  theme_minimal() +
  labs(
    x = "Temperature (Fahrenheit)",
    y = "Velocity (in Vacuum)"
  )
```

**Observations**:

-   After constructing a VelocityVacuum vs. Temp graph, I found that the data between the two sets had some similarities. Similarly to the VelocityVacuum vs. Date graph, the graphs generally took the same shape of when they would rise and fall. That being said, the values themselves were quite different. For example at 90 degrees, the means between the graphs span the entire uncertainty range. Between the range of around 73 and 80 degrees, both sets of data stay within the range of uncertainty. This graph doesn't clearly show us correlation, there is not a clear pattern with how the independent variable affects the dependent variable.
-   After constructing a VelocityVacuum vs. Distinctness graph, I found that the median varies the most from the actual velocity in a vacuum when the distinctness is rated 1. The simulated results actually found that a distinctness of 1 is closest to the actual velocity out of the three levels. From both sets of data, the medians don't stray far from the actual velocity in a vacuum with the exception of the real data with a distinctness of 1.

From these graphs, it seems notable that the lower distinctness deviated from the actual velocity in a vacuum than the other distinctness values. The simulated data that would theoretically get closer to the actual velocity accounted for a lower distinctness being close to the actual velocity. Since this doesn't align with real results, this could be a reason that Michelson's data was off, particularly the uncertainty he expected. Overall, this data lacks many obvious patterns that would point to a reasonable answer for where Michelson is off, but it would not be unreasonable to think that there were errors in his procedure that led to the results he got.

## Bibliography

-   [1] Michelson, [Experimental Determination of the Velocity of Light](https://play.google.com/books/reader?id=343nAAAAMAAJ&hl=en&pg=GBS.PA115) (1880)
-   [2] Henrion and Fischhoff, [Assessing Uncertainty in Physical Constants](https://www.cmu.edu/epp/people/faculty/research/Fischoff-Henrion-Assessing%20uncertainty%20in%20physical%20constants.pdf) (1986)
-   [3] BYU video about a [Fizeau-Foucault apparatus](https://www.youtube.com/watch?v=Ik5ORaaeaME), similar to what Michelson used.
